### 2021-06-18 papers/thesis/full/ch1-introduction/introduction.tex
minor edits after readthrough (typos, repetition of accronym defs, etc.)

Spherical harmonics are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere and form a basis for expanding functions defined on the sphere) and as a result are widely used in many scientific fields for solving PDEs including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory (e.g. \cite{varshalovich1988quantum}), biochemistry (e.g.\cite{parimal2014application, basko1998application}), geosciences (e.g. \cite{fletcher2017data, hollerbach2013parity}) and meteorology (e.g. \cite{evans1998spherical, rubinstein2015scalar, wedi2013fast, ecmwf2020scalability, courtier1998ecmwf, silberman1954planetary, gottlieb1977numerical}). Spectral methods on the sphere involving spherical harmonics have been used for over 60 years \cite{silberman1954planetary}. Notably, spherical harmonics are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}.

@@

\bsrefchapter{CHAPTER:sphericalharmonics} of this thesis provides an introduction to sparse spectral methods via the spherical harmonics on the whole sphere surface. Here, we think of the spherical harmonics in a non-traditional way and write them as a group of multidimensional orthogonal polynomials in $(x,y,z)$ as opposed to functions of spherical coordinates. We present an OP framework for how the spherical harmonics can be used to expand functions defined on the sphere as multidimensional polynomials in $x, y, z$, and how differential operators can be applied as banded-block-banded matrix operators to coefficient vectors for a function's expansion. Further, we demonstrate how the vector spherical harmonics can be used as an orthogonal basis for vector valued functions lying in the \enquote{tangent bundle} of the sphere, and thus how one can additionally derive gradient and divergence operators.

>>>>

Spherical harmonics (SHs) are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere and form a basis for expanding functions defined on the sphere) and as a result are widely used in many scientific fields for solving PDEs including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory (e.g. \cite{varshalovich1988quantum}), biochemistry (e.g.\cite{parimal2014application, basko1998application}), geosciences (e.g. \cite{fletcher2017data, hollerbach2013parity}) and meteorology (e.g. \cite{evans1998spherical, rubinstein2015scalar, wedi2013fast, ecmwf2020scalability, courtier1998ecmwf, silberman1954planetary, gottlieb1977numerical}). Spectral methods on the sphere involving spherical harmonics have been used for over 60 years \cite{silberman1954planetary}. Notably, spherical harmonics are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}.

@@

\bsrefchapter{CHAPTER:sphericalharmonics} of this thesis provides an introduction to sparse spectral methods via the spherical harmonics on the whole sphere surface. Here, we think of the spherical harmonics in a non-traditional way and write them as a group of multidimensional orthogonal polynomials in $(x,y,z)$ as opposed to functions of spherical coordinates. We present an OP framework for how the spherical harmonics can be used to expand functions defined on the sphere as multidimensional polynomials in $x, y, z$, and how differential operators can be applied as banded-block-banded matrix operators to coefficient vectors for a function's expansion. Further, we demonstrate how the vector spherical harmonics (VSHs) can be used as an orthogonal basis for vector valued functions lying in the \enquote{tangent bundle} of the sphere, and thus how one can additionally derive gradient and divergence operators.
